Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-10-14 manu-draft.qmd
FEE formatting and submission

---
format: 
  docx:
    reference-doc: my_styles.docx
title: "Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
shorttitle: "Tampa Bay watershed habitat restoration planning"
bibliography: refs.bib
csl: frontiers-in-ecology-and-evolution.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Douglas E Robison:
      email: DRobison@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Gary E Raulerson:
      email: gmangrove@gmail.com
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      institute: [TBEP]
      correspondence: false
  - Justin Saarinen:
      email: justinsaarinen@gmail.com 
      institute: [ESA]
      correspondence: false
  - Christine Sciarrino:
      email: CSciarrino@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Edward T Sherwood:
      email: esherwood@tbep.org 
      institute: [TBEP]
      correspondence: false
  - David A Tomasko:
      email: dave@sarasotabay.org
      institute: [SBEP]
      correspondence: false
institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - ESA: Environmental Science Associates, Tampa, Florida 33609 USA
  - SBEP: Sarasota Bay Estuary Program, Sarasota, Florida 34236 USA

@@

## Abstract

@@

The health of estuarine systems and coastal habitats is tightly linked to land use and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19]. Management of coastal habitats requires an understanding of how stressors influence these resources, while balancing competing societal uses.

@@

In this paper, we describe an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories. The outcomes are spatially-specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow. The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify opportunities. Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain the targets and goals described above. Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given underlying soil conditions, expected land use change, and sea-level rise.

@@

Priorities for comprehensive, watershed-wide habitat restoration should be informed by current assessments and what is possible to achieve in the future. These priorities are necessary given anticipated impacts of land development and climate change, while also considering competing societal interests for use of the environment and limited resources for land acquisition and restoration. Our approach balances these tradeoffs by identifying targets and goals that are informed by current extent, past trends, and realistic effort from past projects. Further, spatially-explicit locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development). This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate an ideal historical baseline. Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to assess progress, and 2) can be used in other locations with similar needs for identifying restoration priorities.

@@

Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region. The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed. Long-term datasets describing land use and cover and the extent of subtidal habitats were necessary to categorize current extent and past trends. Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline). Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas. Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets. Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets, which could expose additional opportunities in urban or suburban areas (e.g., improving stormwater infrastructure) or differential changes within native habitat classes (e.g., interspecific differences in mangrove colonization). Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types. For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities. Tracking progress towards these habitat targets and goals is heavily reliant on regular updates to these datasets, as well as routine land use and cover map updates.

@@

## Acknowledgments

@@

## Figures

![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA. The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}

![Spatial analysis workflow used to identify opportunity areas (existing conservation native, proposed conservation native, existing conservation restorable, proposed conservation restorable, reservation native, reservation restorable) and restoration potential (coastal uplands, uplands, freshwater wetlands, tidal wetlands) in the Tampa Bay watershed. Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.  The approach was applied to both the intertidal and supratidal strata of the watershed.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}

![Change analysis of habitat categories in the subtidal strata of the Tampa Bay watershed. The left column shows relative areas in 1988 and the right column shows relative areas in 2018 for each habitat category, where the bar heights are proportional to extents in each year. The grey lines show the proportional change in area of each habitat category between the years.](figs/subtchgdatalluv.png){#fig-subtchgdatalluv fig-align="center"}

![Change analysis of habitat categories in the intertidal and supratidal strata of the Tampa Bay watershed. The left column shows relative areas in 1990 and the right column shows relative areas in 2017 for each habitat category, where the bar heights are proportional to extents in each year. The grey lines show the proportional change in area of each habitat category between the years.](figs/chgdatalluv.png){#fig-chgdatalluv fig-align="center"}

![Opportunity areas for habitat restoration in the Tampa Bay watershed. Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservation opportunities. Each category is also grouped into native and restorable habitats. The outline is the Tampa Bay watershed.](figs/oppmap.png){#fig-oppmap fig-align="center"}

![Habitat restoration potential in the Tampa Bay watershed. Areas are identified as those where habitat restoration could target the four identified categories as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands. Categories are based on the opportunity areas, soil types, and coastal boundaries. The outline is the Tampa Bay watershed.](figs/restmap.png){#fig-restmap fig-align="center"}

@@

## Tables

@@

## References

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
shorttitle: "Tampa Bay watershed habitat restoration planning"
bibliography: refs.bib
csl: frontiers-in-ecology-and-evolution.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Douglas E Robison:
      email: DRobison@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Gary E Raulerson:
      email: gmangrove@gmail.com
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      institute: [TBEP]
      correspondence: false
  - Justin Saarinen:
      email: justinsaarinen@gmail.com 
      institute: [ESA]
      correspondence: false
  - Christine Sciarrino:
      email: CSciarrino@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Edward T Sherwood:
      email: esherwood@tbep.org 
      institute: [TBEP]
      correspondence: false
  - David A Tomasko:
      email: dave@sarasotabay.org
      institute: [SBEP]
      correspondence: false
institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - ESA: Environmental Science Associates, Tampa, Florida 33609 USA
  - SBEP: Sarasota Bay Estuary Program, Sarasota, Florida 34236 USA

@@

## Abstract {.unnumbered}

@@

The health of estuarine systems and coastal habitats is tightly linked to land use and management of the watershed [@Yoskowitz15]. Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18]. Pervasive stressors can negatively impact the services provided by these habitats. Anthropogenic land development can remove or substantially modify existing habitats and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14]. Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19]. Management of coastal habitats requires an understanding of how stressors influence these resources, while balancing competing societal uses.

@@

In this paper, we describe an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories. The outcomes are spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Spatially explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow. The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify opportunities. Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain the targets and goals described above. Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given underlying soil conditions, expected land use change, and sea-level rise.

@@

Priorities for comprehensive, watershed-wide habitat restoration should be informed by current assessments and what is possible to achieve in the future. These priorities are necessary given anticipated impacts of land development and climate change, while also considering competing societal interests for use of the environment and limited resources for land acquisition and restoration. Our approach balances these tradeoffs by identifying targets and goals that are informed by current extent, past trends, and realistic effort from past projects. Further, spatially explicit locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development). This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate an ideal historical baseline. Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to assess progress, and 2) can be used in other locations with similar needs for identifying restoration priorities.

@@

Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region. The workflow for identifying priorities required detailed and spatially explicit datasets specific to the Tampa Bay watershed. Long-term datasets describing land use and cover and the extent of subtidal habitats were necessary to categorize current extent and past trends. Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline). Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas. Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets. Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets, which could expose additional opportunities in urban or suburban areas (e.g., improving stormwater infrastructure) or differential changes within native habitat classes (e.g., interspecific differences in mangrove colonization). Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types. For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities. Tracking progress towards these habitat targets and goals is heavily reliant on regular updates to these datasets, as well as routine land use and cover map updates.

@@

## Acknowledgments {.unnumbered}

@@

## Figures {.unnumbered}

![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA. The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}

![Spatial analysis workflow used to identify opportunity areas (existing conservation native, proposed conservation native, existing conservation restorable, proposed conservation restorable, reservation native, reservation restorable) and restoration potential (coastal uplands, uplands, freshwater wetlands, tidal wetlands) in the Tampa Bay watershed. Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.  The approach was applied to both the intertidal and supratidal strata of the watershed.](figs/gisworkflow.jpg){#fig-gisworkflow fig-align="center"}

![Change analysis of habitat categories in the subtidal strata of the Tampa Bay watershed. The left column shows relative areas in 1988 and the right column shows relative areas in 2018 for each habitat category, where the bar heights are proportional to extents in each year. The grey lines show the proportional change in area of each habitat category between the years.](figs/subtchgdatalluv.jpg){#fig-subtchgdatalluv fig-align="center"}

![Change analysis of habitat categories in the intertidal and supratidal strata of the Tampa Bay watershed. The left column shows relative areas in 1990 and the right column shows relative areas in 2017 for each habitat category, where the bar heights are proportional to extents in each year. The grey lines show the proportional change in area of each habitat category between the years.](figs/chgdatalluv.jpg){#fig-chgdatalluv fig-align="center"}

![Opportunity areas for habitat restoration in the Tampa Bay watershed. Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservation opportunities. Each category is also grouped into native and restorable habitats. The outline is the Tampa Bay watershed.](figs/oppmap.jpg){#fig-oppmap fig-align="center"}

![Habitat restoration potential in the Tampa Bay watershed. Areas are identified as those where habitat restoration could target the four identified categories as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands. Categories are based on the opportunity areas, soil types, and coastal boundaries. The outline is the Tampa Bay watershed.](figs/restmap.jpg){#fig-restmap fig-align="center"}

@@

## Tables {.unnumbered}

@@

## References {.unnumbered}
